Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-03 report/report.Rmd
.

---
title: "Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas"
author: "Léon Kia Faro"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    number_sections: true
    toc: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    df_print: paged
fontsize: 11pt
geometry: margin=1in
linkcolor: blue
---

@@

Estimating the joint density $p(x_1,\dots,x_K)$ of a multivariate random vector $x=(x_1,\ldots,x_K)$ is a core problem in statistical modeling and unsupervised learning. By learning $p(x)$ from data, we enable a range of tasks including probabilistic inference, anomaly detection, data imputation, and generative modeling. However, high-dimensional density estimation is challenging due to the curse of dimensionality and complex dependency structures. This thesis investigates three modern approaches to multivariate density estimation – **Triangular Transport Maps (TTM)**, **Transformation Random Forests (TRTF)**, and **Copula Models** – comparing their methodologies and performance. We focus on how each method transforms or models a complex joint distribution by leveraging simpler components, and we evaluate them on benchmark data.

A unifying perspective for density estimation is **measure transport**, where one transforms a complicated **target distribution** $\pi$ on $\mathbb{R}^K$ into a simpler **reference distribution** $\eta$ (often a standard Gaussian or uniform). If $S:\mathbb{R}^K \to \mathbb{R}^K$ is an invertible transport map such that $z = S(x)$ has distribution $\eta$ when $x \sim \pi$, then we can estimate $\pi$ via learning $S$ and use $S^{-1}$ for sampling from $\pi$. **Normalizing flows** implement this idea using deep neural networks: they construct $S$ as a composition of many invertible transformations with tractable Jacobians. *Triangular transport maps*, in contrast, impose a structured, *monotonic* form on $S$ that offers theoretical and practical advantages. Meanwhile, **transformation forests** (TRTF) take a nonparametric ensemble approach, using tree-based models to estimate conditional distributions. Finally, **copula models** provide a statistically interpretable way to construct multivariate densities by separating marginal distributions from the dependency structure (the copula). In this work, we formulate each approach in a common framework, strictly adopting the notation and terminology of a recent tutorial on triangular transport maps. We then empirically compare their performance on a real dataset and discuss theoretical connections between the methods.

@@

We address the following research questions:

- **Q1: Performance** – How do TTM, TRTF, and copula models compare in terms of goodness-of-fit for multivariate density estimation? We will evaluate models on held-out test data via log-likelihood and other metrics to determine which approach best captures the true distribution of complex, high-dimensional data.

- **Q2: Trade-offs** – What are the trade-offs in complexity, interpretability, and computational efficiency between the methods? For instance, triangular maps offer *exact* likelihood evaluation and have interpretable structure, while normalizing flows and forests might be more flexible but less transparent. We also ask if simpler parametric copulas suffer when assumptions are violated.

@@

We outline the methodology of each approach. We begin with triangular transport maps (a type of normalizing flow with a structured form), then describe transformation random forests, copula models, and finally define baseline oracle models. Throughout, we let $\pi(x)$ denote the unknown target density of interest (on $\mathbb{R}^K$) and $\eta(z)$ a reference density (typically a standard multivariate normal or product of uniforms). All methods ultimately seek to estimate $\pi$ by either constructing a map $S$ such that $S(X)\sim \eta$ for $X\sim \pi$, or by directly modeling $\pi$ via factorization into lower-dimensional components.

@@

We adopt the transport view: learn an invertible map \(S:\mathbb{R}^K\to\mathbb{R}^K\) that pushes the target \(\pi\) to a reference \(\eta\) (typically \(\mathcal N(0,I)\)). A map \(S\) is **triangular** if
\[
S(x)=(S_1(x_1),\ S_2(x_{1:2}),\ \dots,\ S_K(x_{1:K})),
\]
and each \(S_k\) is **strictly increasing in its last argument** \(x_k\). This structure implies a factorized Jacobian
\[
\det \nabla S(x) \;=\; \prod_{k=1}^K \partial_{x_k} S_k(x_{1:k}),
\]
which makes change‑of‑variables numerically stable.

@@

### Parameterizations used in this thesis
We study three monotone triangular families (all evaluated on standardized inputs \(u\)):

@@

## Transformation Random Forests

@@

## Exact likelihoods under triangular maps and TRTF (new)

@@

Copulas decouple marginals from dependence: with $u_i = F_{X_i}(x_i)$ and copula density $c(u)$, the joint density factorizes as $\pi(x) = c(u)\prod_i f_{X_i}(x_i)$. We consider a Gaussian copula with empirical marginals (semiparametric) and a smoothed empirical copula (nonparametric). These highlight the cost of misspecified dependence (parametric) versus the data demands of flexible nonparametric dependence estimation.

@@

## Data and Preprocessing

For MiniBooNE we follow standard preprocessing: train-only standardization, removal of near-constant attributes, and pruning highly correlated features (to avoid trivial ridges). Established train/val/test splits are used to ensure comparability with prior work.

## Log-Likelihood Estimation

Primary metric is *test* log-likelihood in nats. TTM and copulas allow exact evaluation via change-of-variables and copula factorization, respectively. TRTF evaluates the sum of conditional pdf logs, using the local parametric family from each forest leaf. We also compute calibration checks (marginal QQ) and dependence diagnostics (rank correlations) as secondary assessments.

## Fairness and Robustness

All methods share the same splits, seeds, and comparable complexity; each result is averaged over multiple seeds with standard errors reported. We apply paired tests or overlap of ±2·SE to gauge practical significance differences.

@@

|  # | model               | mean\_joint\_nll | per\_dim\_nll\_1 | per\_dim\_nll\_2 |
| -: | ------------------- | ---------------- | ---------------- | ---------------- |
|  1 | True\_uncond. | 1.37 ± 0.11      | 0.69             | 0.69             |
|  2 | True\_cond.   | 0.70 ± 0.12      | 0.35             | 0.35             |
|  3 | TRTF                | 1.83 ± 0.14      | 1.25             | 0.57             |
|  4 | TTM\_marginal                 | 2.04 ± 0.12      | 1.29             | 0.75             |
|  5 | TTM\_sep            | 1.92 ± 0.14      | 1.29             | 0.64             |
|  6 | TTM\_cross          | 1.22 ± 0.20      | 0.92             | 0.29             |
|  7 | Copula\_np          | 0.87 ± 0.16      | 0.76             | 0.11             |

@@

|Dim | Distribution | True (marginal) | True (joint) | Random Forest | Marginal Map | Separable Map | Cross-Term Map |
|--: | :--------- | :-------------: | :----------: | :-----------: | :----------: | :-----------: | :------------: |
| 1 | norm       |   1.46 ± 0.26   |  1.41 ± 0.29 |  1.48 ± 0.24  |  1.49 ± 0.20 |  1.46 ± 0.26  |   1.46 ± 0.25  |
| 2 | exp        |   1.55 ± 0.46   |  1.38 ± 0.65 |  2.54 ± 0.75  |  3.30 ± 0.01 |  1.75 ± 0.70  |   2.58 ± 0.03  |
| 3 | beta       |   −0.46 ± 0.63  | −0.63 ± 1.00 |  −0.14 ± 0.34 |  0.40 ± 0.17 |  0.28 ± 0.70  |   0.39 ± 0.23  |
| 4 | gamma      |   2.21 ± 1.11   |  2.07 ± 0.80 |  2.22 ± 1.08  |  2.78 ± 0.77 |  2.97 ± 1.45  |   3.00 ± 1.46  |
| k | SUM        |   4.75 ± 1.10   |  4.23 ± 1.03 |  6.10 ± 1.61  |  7.97 ± 0.94 |  6.47 ± 1.92  |   7.43 ± 1.66  |

@@

| **Model**             | **Miniboone ** |
|-----------------------|------------------------------|
| Gaussian (indep. baseline)  | $-37.24 \pm 1.07$ |
| MADE (ACN)            | $-15.59 \pm 0.50$ |
| MADE MoG              | $-12.27 \pm 0.47$ |
| Real NVP (5-layer)    | $-13.55 \pm 0.49$ |
| Real NVP (10-layer)   | $-13.84 \pm 0.52$ |
| MAF (5-layer)         | $-11.75 \pm 0.44$ |
| MAF (10-layer)        | $-12.24 \pm 0.45$ |
| MAF MoG (5-layer)     | **$-11.68 \pm 0.44$** |
| **TRTF (Transformation Forest)** | $-29.88 \pm 0.02$ *(ours)* |

@@

@@

@@

# Theoretical Links between TRTF and TTM

Let $\hat{F}_k$ be the TRTF estimate of $F_{X_k|X_{1:k-1}}$. Define $\hat{S}_k(x_{1:k})=\hat{F}_k(x_k\mid x_{1:k-1})$. Then $\hat{\mathbf{S}}(x)=(\hat{S}_1,\dots,\hat{S}_K)$ is triangular and monotone in the last argument by construction, pushing the empirical distribution toward $\mathrm{Unif}(0,1)^K$. In the limit of infinite data and perfect conditional estimation, $\hat{\mathbf{S}}$ converges to the KR map. Thus, TRTF can be viewed as a nonparametric learner of triangular transport, providing a principled bridge between tree ensembles and measure transport.

@@

**Predict.** For each \(x\) (row-wise):
1. \(u=(x-\mu)\oslash\sigma\).
2. For each \(k\):
   \[
   \pi_{k,\mathrm{std}}(u_k)=\sigma_k\,\pi_k(\mu_k+\sigma_k u_k),\qquad
   \mathrm{LD}_k=\log\pi_{k,\mathrm{std}}(u_k)-\log\sigma_k=\log\pi_k(x_k).
   \]
3. Output \(\mathrm{LD}=(\mathrm{LD}_k)_{k=1}^K\), \(L=\sum_k \mathrm{LD}_k\).

*(Unit test: if \(\pi_k=\mathcal{N}(0,1)\) and \(\mu=0,\sigma=1\), then \(\mathrm{LD}_k=-\tfrac12x_k^2-\tfrac12\log(2\pi)\).)*

@@

**Training objective (maps from samples; Gaussian reference).**
\(
\min_{\{a_k,b_k>0\}}\ \sum_{i\in\mathrm{tr}}\sum_{k=1}^K
\Big(\tfrac12 S_k(u_{ik})^2-\log b_k\Big).
\)

@@

**Parameterization (on standardized \(u\)).**
\(
S_k(u_{1:k})=g_k(u_{1:k-1})+f_k(u_k),\qquad f_k'\!(u_k)>0.
\)
Typical basis: \(g_k(u_{1:k-1})=\sum_j c^{\text{non}}_{k,j}\,\psi^{\text{non}}_{k,j}(u_{1:k-1})\),  
\(f_k(u_k)=\sum_j c^{\text{mon}}_{k,j}\,\psi^{\text{mon}}_{k,j}(u_k)\) with \(\psi^{\text{mon}}\) monotone (e.g. iRBF/edge terms).

**Training objective (decouples by \(k\)).**
\(
\min_{\{c^{\text{non}}_k,c^{\text{mon}}_k\}}\ J_k(c^{\text{non}}_k,c^{\text{mon}}_k)
=\sum_{i\in\mathrm{tr}}\Big(\tfrac12 S_k(u_{i,1:k})^2-\log f_k'(u_{ik})\Big).
\)
(Option: eliminate \(c^{\text{non}}_k\) in closed form given \(c^{\text{mon}}_k\) via normal equations; then solve a convex box-constrained problem in \(c^{\text{mon}}_k\ge 0\).)

@@

**Training objective (per \(k\)).**
\(
\min_{\{c^{\text{non}}_k,c^{\text{cr}}_k\}}\ 
J_k=\sum_{i\in\mathrm{tr}}\Big(\tfrac12 S_k(u_{i,1:k})^2-\log\partial_{u_k}S_k(u_{i,1:k})\Big),
\quad
\partial_{u_k}S_k=r\big(h_k(u_{ik},u_{i,1:k-1})\big).
\)
(Compute the integral by 1D quadrature; enforce stability by clipping \(h_k\) to \([{-}H,H]\) during training/inference.)

**Predict.** For each \(x\):
\(
u=(x-\mu)\oslash\sigma,\quad z_k=S_k(u_{1:k}),\quad 
\mathrm{LD}_k=-\tfrac12 z_k^2-\tfrac12\log(2\pi)+\log r\!\big(h_k(u_k,u_{1:k-1})\big)-\log\sigma_k,\quad
L=\sum_k\mathrm{LD}_k.
\)

## Appendix B — Acceptance checks (all models; required)

>>>>

---
title: "Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas"
author: "Léon Kia Faro"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    number_sections: true
    toc: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
df_print: paged
fontsize: 11pt
geometry: margin=1in
linkcolor: blue
header-includes:
  - \usepackage{amsmath,amssymb,mathtools}
  - \usepackage{algorithm}
  - \usepackage{algpseudocode}   % provides \begin{algorithmic}, \State, \Function, \Comment, \Statex
  - \floatname{algorithm}{Algorithm}
  - \renewcommand{\algorithmicrequire}{\textbf{Input:}}
  - \renewcommand{\algorithmicensure}{\textbf{Output:}}
---

@@

Estimating the joint density \(p(x_1,\dots,x_K)\) of a multivariate random vector \(x=(x_1,\ldots,x_K)\) is a core problem in statistical modeling and unsupervised learning. By learning \(p(x)\) from data, we enable a range of tasks including probabilistic inference, anomaly detection, data imputation, and generative modeling. However, high-dimensional density estimation is challenging due to the curse of dimensionality and complex dependency structures.

This thesis investigates three modern approaches to multivariate density estimation – **Triangular Transport Maps (TTM)**, **Transformation Random Forests (TRTF)**, and **Copula Models** – comparing their methodologies and performance. We focus on how each method transforms or models a complex joint distribution by leveraging simpler components, and we evaluate them on benchmark data.

A unifying perspective for density estimation is **measure transport**, where one transforms a complicated **target distribution** \(\pi\) on \(\mathbb{R}^K\) into a simpler **reference distribution** \(\eta\) (often a standard Gaussian or uniform). If \(S:\mathbb{R}^K \to \mathbb{R}^K\) is an invertible transport map such that \(z = S(x)\) has distribution \(\eta\) when \(x \sim \pi\), then we can estimate \(\pi\) via learning \(S\) and use \(S^{-1}\) for sampling from \(\pi\). **Normalizing flows** implement this idea using deep neural networks: they construct \(S\) as a composition of many invertible transformations with tractable Jacobians. *Triangular transport maps*, in contrast, impose a structured, *monotonic* form on \(S\) that offers theoretical and practical advantages. Meanwhile, **transformation forests** (TRTF) take a nonparametric ensemble approach, using tree-based models to estimate conditional distributions. Finally, **copula models** provide a statistically interpretable way to construct multivariate densities by separating marginal distributions from the dependency structure (the copula). In this work, we formulate each approach in a common framework, strictly adopting the notation and terminology of a recent tutorial on triangular transport maps. We then empirically compare their performance on a real dataset and discuss theoretical connections between the methods.

## Contributions

- **Unified formulation:** We place TTM, TRTF, and copulas under a single change-of-variables and conditional-factorization framework with **train-only standardization**, **per-dimension log-density accounting in nats**, and a common **Predict-API** (matrix `logdensity_by_dim`; vector `logdensity` as row sums).
- **Theory bridge:** We provide a concise derivation showing that **TRTF induces a separable triangular map**; the TRTF joint likelihood equals the sum of conditional log-pdfs (plus the single standardization Jacobian).
- **Evaluation protocol:** Deterministic seeds, identical splits and preprocessing, acceptance checks (shapes, constants, stability), and reporting of **joint NLL** and **per-dimension NLL** with standard errors.
- **Comparative study:** Experiments on synthetic settings (Half-Moon; 4D conditional generator) and real data (MINIBOONE), highlighting strengths, weaknesses, and trade-offs of each class.

@@

- **Q1: Performance** — How do TTM, TRTF, and copula models compare in terms of goodness-of-fit for multivariate density estimation? We evaluate models on held-out test data via log-likelihood and per-dimension contributions.
- **Q2: Trade-offs** — What are the trade-offs among complexity, interpretability, and computational efficiency? Triangular maps yield exact likelihoods and interpretable Jacobians; forests and flows may be more flexible but less transparent.
- **Q3: Theory** — In what precise sense does TRTF realize a triangular transport, and when might cross-term transports exceed separable maps?

@@

We outline the methodology for each approach. We let \(\pi(x)\) denote the unknown target density (on \(\mathbb{R}^K\)) and \(\eta(z)\) a reference (typically \(\mathcal N(0,I)\) or product uniforms). All methods estimate \(\pi\) either by constructing \(S\) with \(S(X)\sim\eta\) for \(X\sim\pi\) or by factorizing \(\pi\) into lower-dimensional components.

## Common setup (summary; see Appendix A for full pseudoalgorithms)

- **Train-only standardization.** For \(k=1,\dots,K\),
  \[
  u_k \,=\, \frac{x_k-\mu_k}{\sigma_k},\qquad \sigma_k>0,
  \]
  where \((\mu_k,\sigma_k)\) are computed on train and persisted to val/test. This preprocessing contributes **exactly once per dimension** the Jacobian term \(-\log\sigma_k\) to the log-density.
- **Predict-API invariants.** `predict(., "logdensity_by_dim") \in \mathbb{R}^{N\times K}`; `predict(., "logdensity") \in \mathbb{R}^N` with row sums. Units are **nats**; no NA/Inf.
- **Determinism.** A global seed fixes splits, training, and evaluation; repeated runs match up to machine precision.

@@

We adopt the transport view: learn an invertible map \(S:\mathbb{R}^K\to\mathbb{R}^K\) that pushes \(\pi\) to \(\eta\) (typically \(\mathcal N(0,I)\)). A map \(S\) is **triangular** if
\[
S(x)=(S_1(x_1),\ S_2(x_{1:2}),\ \dots,\ S_K(x_{1:K})),
\]
and each \(S_k\) is **strictly increasing in its last argument** \(x_k\). This structure implies a factorized Jacobian
\[
\det \nabla S(x) \;=\; \prod_{k=1}^K \partial_{x_k} S_k(x_{1:k}),
\]
which makes change‑of‑variables numerically stable.

@@

### Parameterizations used in this thesis
We study three monotone triangular families (evaluated on standardized inputs \(u\)):

@@

## Transformation Random Forests (TRTF)

@@

## Exact likelihoods under triangular maps and TRTF

@@

Copulas decouple marginals from dependence: with \(u_i = F_{X_i}(x_i)\) and copula density \(c(u)\), the joint density factorizes as \(\pi(x) = c(u)\prod_i f_{X_i}(x_i)\). We consider a Gaussian copula with empirical marginals (semiparametric) and a smoothed empirical copula (nonparametric). These highlight the cost of misspecified dependence (parametric) versus the data demands of flexible nonparametric dependence estimation.

@@

## Data and Preprocessing
For MiniBooNE we follow standard preprocessing: train-only standardization, removal of near-constant attributes, and pruning highly correlated features (to avoid trivial ridges). Established train/val/test splits are used to ensure comparability with prior work.

## Log-Likelihood Estimation
Primary metric is *test* log-likelihood in nats. TTM and copulas allow exact evaluation via change-of-variables and copula factorization, respectively. TRTF evaluates the sum of conditional pdf logs, using the local parametric family from each forest leaf. We also compute calibration checks (marginal QQ) and dependence diagnostics (rank correlations) as secondary assessments.

## Fairness and Robustness
All methods share the same splits, seeds, and comparable complexity; each result is averaged over multiple seeds with standard errors reported. We apply paired tests or overlap of \(\pm 2\cdot\mathrm{SE}\) to gauge practical significance differences.

@@

|  # | model               | mean_joint_nll | per_dim_nll_1 | per_dim_nll_2 |
| -: | ------------------- | -------------- | ------------- | ------------- |
|  1 | True_uncond.        | 1.37 ± 0.11    | 0.69          | 0.69          |
|  2 | True_cond.          | 0.70 ± 0.12    | 0.35          | 0.35          |
|  3 | TRTF                | 1.83 ± 0.14    | 1.25          | 0.57          |
|  4 | TTM_marginal        | 2.04 ± 0.12    | 1.29          | 0.75          |
|  5 | TTM_sep             | 1.92 ± 0.14    | 1.29          | 0.64          |
|  6 | TTM_cross           | 1.22 ± 0.20    | 0.92          | 0.29          |
|  7 | Copula_np           | 0.87 ± 0.16    | 0.76          | 0.11          |

@@

|Dim | Distribution | True (marginal) | True (joint) | Random Forest | Marginal Map | Separable Map | Cross-Term Map |
|--: | :----------- | :-------------: | :----------: | :-----------: | :----------: | :-----------: | :------------: |
| 1  | norm         |   1.46 ± 0.26   |  1.41 ± 0.29 |  1.48 ± 0.24  |  1.49 ± 0.20 |  1.46 ± 0.26  |   1.46 ± 0.25  |
| 2  | exp          |   1.55 ± 0.46   |  1.38 ± 0.65 |  2.54 ± 0.75  |  3.30 ± 0.01 |  1.75 ± 0.70  |   2.58 ± 0.03  |
| 3  | beta         |   −0.46 ± 0.63  | −0.63 ± 1.00 |  −0.14 ± 0.34 |  0.40 ± 0.17 |  0.28 ± 0.70  |   0.39 ± 0.23  |
| 4  | gamma        |   2.21 ± 1.11   |  2.07 ± 0.80 |  2.22 ± 1.08  |  2.78 ± 0.77 |  2.97 ± 1.45  |   3.00 ± 1.46  |
| k  | SUM          |   4.75 ± 1.10   |  4.23 ± 1.03 |  6.10 ± 1.61  |  7.97 ± 0.94 |  6.47 ± 1.92  |   7.43 ± 1.66  |

@@

| **Model**                     | **Miniboone**           |
|------------------------------|-------------------------|
| Gaussian (indep. baseline)   | \(-37.24 \pm 1.07\)     |
| MADE (ACN)                   | \(-15.59 \pm 0.50\)     |
| MADE MoG                     | \(-12.27 \pm 0.47\)     |
| Real NVP (5-layer)           | \(-13.55 \pm 0.49\)     |
| Real NVP (10-layer)          | \(-13.84 \pm 0.52\)     |
| MAF (5-layer)                | \(-11.75 \pm 0.44\)     |
| MAF (10-layer)               | \(-12.24 \pm 0.45\)     |
| MAF MoG (5-layer)            | **\(-11.68 \pm 0.44\)** |
| **TRTF (Transformation Forest)** | \(-29.88 \pm 0.02\) *(ours)* |

@@

(*Add figures/tables with per-dimension NLL and calibration plots; the exact model likelihood allows comprehensive tail and PIT diagnostics.*)

@@

(*Report both parametric Gaussian and nonparametric copulas; include marginal calibration and tail dependence measures.*)

@@

(*For simulations, include oracle baselines to contextualize achievable performance.*)

# Discussion and Conclusion

**Methodological takeaway.** TRTF realizes a **separable** triangular transport and thus provides exact, normalized likelihoods via the sum of conditional log-pdfs. Separable TTM and TRTF excel on problems dominated by conditional shifts and monotone shape changes. Cross-term triangular transports can capture complex, localized interactions beyond separable structure but require careful numerical stabilization.

**Comparative view.** Parametric copulas are strong under elliptical dependence and weak otherwise; nonparametric copulas can adapt to local/tail asymmetries at the cost of higher data and smoothing choices. Neural flows (deep normalizing flows) remain the most flexible among the classes studied, but pay in interpretability and hyperparameter complexity.

**Practical guidance.** Start with separable TTM/TRTF for structured tabular problems with moderate \(K\) and a premium on interpretability and exact likelihoods. Escalate to cross-term TTM or neural flows when interactions/tails overwhelm separable structure, and to nonparametric copulas when marginal calibration is critical and data are abundant.

@@

**Predict.** For each \(x\) (row-wise): 1. \(u=(x-\mu)\oslash\sigma\). 2. For each \(k\):
\[
\pi_{k,\mathrm{std}}(u_k)=\sigma_k\,\pi_k(\mu_k+\sigma_k u_k),\qquad
\mathrm{LD}_k=\log\pi_{k,\mathrm{std}}(u_k)-\log\sigma_k=\log\pi_k(x_k).
\]
3. Output \(\mathrm{LD}=(\mathrm{LD}_k)_{k=1}^K\), \(L=\sum_k \mathrm{LD}_k\).
*(Unit test: if \(\pi_k=\mathcal{N}(0,1)\) and \(\mu=0,\sigma=1\), then \(\mathrm{LD}_k=-\tfrac12x_k^2-\tfrac12\log(2\pi)\).)*

@@

**Training objective (maps from samples; Gaussian reference).**
\[
\min_{\{a_k,b_k>0\}} \ \sum_{i\in\mathrm{tr}}\sum_{k=1}^K \Big(\tfrac12 S_k(u_{ik})^2-\log b_k\Big).
\]

@@

**Parameterization (on standardized \(u\)).**
\( S_k(u_{1:k})=g_k(u_{1:k-1})+f_k(u_k),\qquad f_k'\!(u_k)>0. \)
Typical basis: \(g_k(u_{1:k-1})=\sum_j c^{\text{non}}_{k,j}\,\psi^{\text{non}}_{k,j}(u_{1:k-1})\),
\( f_k(u_k)=\sum_j c^{\text{mon}}_{k,j}\,\psi^{\text{mon}}_{k,j}(u_k)\) with \(\psi^{\text{mon}}\) monotone (e.g. iRBF/edge terms).

**Training objective (decouples by \(k\)).**
\[
\min_{\{c^{\text{non}}_k,c^{\text{mon}}_k\}} \ J_k(c^{\text{non}}_k,c^{\text{mon}}_k)
= \sum_{i\in\mathrm{tr}} \Big(\tfrac12 S_k(u_{i,1:k})^2 - \log f_k'(u_{ik})\Big).
\]
(Option: eliminate \(c^{\text{non}}_k\) in closed form given \(c^{\text{mon}}_k\) via normal equations; then solve a convex box-constrained problem in \(c^{\text{mon}}_k\ge 0\).)

@@

**Training objective (per \(k\)).**
\[
\min_{\{c^{\text{non}}_k,c^{\text{cr}}_k\}} \ J_k
= \sum_{i\in\mathrm{tr}} \Big(\tfrac12 S_k(u_{i,1:k})^2 - \log \partial_{u_k}S_k(u_{i,1:k})\Big), \quad
\partial_{u_k}S_k = r\big(h_k(u_{ik},u_{i,1:k-1})\big).
\]
(Compute the integral by 1D quadrature; enforce stability by clipping \(h_k\) to \([{-}H,H]\) during training/inference.)

**Predict.** For each \(x\):
\( u=(x-\mu)\oslash\sigma,\quad z_k=S_k(u_{1:k}),\quad
 \mathrm{LD}_k=-\tfrac12 z_k^2-\tfrac12\log(2\pi)+\log r\!\big(h_k(u_k,u_{1:k-1})\big)-\log\sigma_k,\quad L=\sum_k\mathrm{LD}_k. \)
# Appendix A — Peudoalgorithms

## 4D Data Gernation

\begin{algorithm}[H]
\caption{\textbf{Triangular Data-Generating Process (DGP) from Configuration}}
\label{alg:dgp-config}
\small
\begin{algorithmic}[1]
\Require Sample size $N\in\mathbb{N}$; dimension $K\in\mathbb{N}$; configuration $\mathrm{cfg}=\{c_k\}_{k=1}^K$ with fields $c_k.\mathrm{distr}\in\{$named 1D family$\}$ and parameter functor $c_k.\mathrm{parm}:\mathbb{R}^{k-1}\!\to\!\mathbb{R}^{m_k}$ (possibly empty); flag $\mathrm{return\_params}\in\{\texttt{TRUE},\texttt{FALSE}\}$.
\Ensure Matrix $X\in\mathbb{R}^{N\times K}$ drawn via the autoregressive/triangular scheme $x_k \sim F_k(\,\cdot\,;\theta_k(x_{1:k-1}))$; optionally a per-dimension parameter history.
\Statex

\State \textbf{Global constants:} numerical floor $\varepsilon\leftarrow 10^{-3}$ for strictly positive parameters.

\Function{Sanitize}{$\theta$} \Comment{\emph{ensure positivity/finite}}
  \ForAll{scalar components $p$ of $\theta$}
     \State \textbf{if} $p\notin(0,\infty)$ \textbf{then} $p\leftarrow \varepsilon$
  \EndFor
  \State \Return $\theta$
\EndFunction

\Function{NormalizeGammaKeys}{$\theta,\,\mathrm{distr}$} \Comment{\emph{compatibility mapping}}
  \If{$\mathrm{distr}=\text{Gamma}$ \textbf{and} $\theta$ has keys \texttt{shape1}, \texttt{shape2}}
     \State $\theta \leftarrow \{\texttt{shape}=\theta[\texttt{shape1}],\ \texttt{scale}=\theta[\texttt{shape2}]\}$
  \EndIf
  \State \Return $\theta$
\EndFunction

\Function{GenerateIIDFromConfig}{$N,\mathrm{cfg},\mathrm{return\_params}$}
  \State Initialize $X\in\mathbb{R}^{N\times K}$ as \texttt{NA}; set column names $(X1,\dots,XK)$
  \If{$\mathrm{return\_params}$}
     \State Initialize $\mathsf{ParamHist}[k]$ as list of length $N$ for each $k=1{:}K$
  \EndIf
  \For{$i=1{:}N$} \Comment{\emph{i.i.d. rows; sequential across dimensions}}
     \For{$k=1{:}K$}
        \State $c \leftarrow \mathrm{cfg}[k]$; \hspace{0.6em} $x_{1:k-1}\leftarrow X_{i,1:(k-1)}$
        \State $\theta_k \leftarrow \begin{cases}
           \varnothing, & \text{if } c.\mathrm{parm}=\varnothing\\
           c.\mathrm{parm}(x_{1:k-1}), & \text{otherwise}
        \end{cases}$
        \State $\theta_k \leftarrow$ \Call{NormalizeGammaKeys}{$\theta_k,\,c.\mathrm{distr}$}
        \State $\theta_k \leftarrow$ \Call{Sanitize}{$\theta_k$}
        \State Draw $X_{i,k} \sim r_{c.\mathrm{distr}}(1;\theta_k)$ \Comment{\emph{one-sample RNG of the named family}}
        \If{$\mathrm{return\_params}$ \textbf{and} $\theta_k\neq\varnothing$}
           \State $\mathsf{ParamHist}[k][i] \leftarrow \theta_k$ \Comment{\emph{store named parameter list for this $(i,k)$}}
        \EndIf
     \EndFor
  \EndFor
  \If{$\mathrm{return\_params}$}
     \State For each $k$: collate non-empty $\mathsf{ParamHist}[k][i]$ across $i$ into a tabular record $\mathsf{Params}[k]$ (row-binded by matching names)
     \State \Return $\{X,\ \mathsf{Params}[1{:}K]\}$
  \Else
     \State \Return $X$
  \EndIf
\EndFunction

\Function{GenSamples}{$G,\mathrm{return\_params}$} \Comment{\emph{compatibility wrapper}}
  \State \Return \Call{GenerateIIDFromConfig}{$G.n,\ G.config,\ \mathrm{return\_params}$}
\EndFunction

\Statex
\State \textbf{Notes.} (i) The triangular structure is enforced by iterating $k=1{:}K$ and letting $\theta_k$ depend only on $x_{1:k-1}$.
(ii) Determinism is controlled by the \emph{caller}'s RNG seed; this routine draws from the current RNG state.
(iii) No standardization is applied at generation time (raw $X$); any train-only standardization and its Jacobian correction are handled downstream during density evaluation.
\end{algorithmic}
\end{algorithm}

## TRTF

\begin{algorithm}[H] % use [H] to place the float exactly here
\caption{\textbf{TRTF (Script-level) — Training, Prediction, and NLL Wrappers}}
\label{alg:trtf-script}
\small
\begin{algorithmic}[1]
\Require Data $X\in\mathbb{R}^{n\times K}$; hyperparameters $(\texttt{ntree},\texttt{minsplit},\texttt{minbucket},\texttt{maxdepth},\texttt{seed})$; $\texttt{cores}$.
\Ensure Model $\mathsf{M}_{\mathrm{core}}$ of class \texttt{mytrtf} with fields $\{\mathsf{ymod}[1{:}K],\ \mathsf{forest}[1{:}(K-1)],\ \mathsf{varimp},\ \texttt{seed}\}$; predictors returning per-dimension and joint log-densities (nats); invariants: finiteness and row-sum equality.

\Function{TrainTRTFCore}{$X,\texttt{ntree},\texttt{minsplit},\texttt{minbucket},\texttt{maxdepth},\texttt{seed},\texttt{cores}$}
  \State \textbf{assert} $X$ numeric $n\times K$; set RNG seed $\texttt{seed}$
  \State $D\leftarrow$ data frame of $X$ with columns $(X1,\dots,XK)$
  \State $\mathsf{ymod}[k]\leftarrow \textsc{BoxCox}(X_k\sim 1;\ D)$ for $k=1{:}K$ \Comment{Normal reference; includes $-\tfrac12\log(2\pi)$ \emph{once}}
  \State $\mathsf{ctrl}\leftarrow \textsc{ctree\_control}(\texttt{minsplit},\texttt{minbucket},\texttt{maxdepth})$
  \For{$k=2{:}K$}
     \State $\mathrm{fm}: X_k \sim X_1+\cdots+X_{k-1}$; $mtry\leftarrow \max\{1,\lfloor (k-1)/2 \rfloor\}$
     \State $\mathsf{forest}[k-1]\leftarrow \textsc{traforest}(\mathsf{ymod}[k],\ \mathrm{fm},\ D;\ \texttt{trace}=\texttt{TRUE},\ \texttt{ntree},\ \texttt{mtry},\ \texttt{mltargs}=\emptyset,\ \texttt{cores},\ \mathsf{ctrl})$
  \EndFor
  \State $\mathsf{varimp}[k-1]\leftarrow \textsc{varimp}(\mathsf{forest}[k-1])$ for $k=2{:}K$
  \State \Return $\mathsf{M}_{\mathrm{core}}=\{\mathsf{ymod}[1{:}K],\ \mathsf{forest}[1{:}(K-1)],\ \mathsf{varimp},\ \texttt{seed}\}$ with class \texttt{mytrtf}
\EndFunction
